---
title: "R Day 2"
author: "Michel LeRoy"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data", "gapminder5.csv")

gapminder <- read.csv(here::here("data", "gapminder5.csv"))
```


## Converting factors to characters
```{r}
str(gapminder)
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## Getting GDP from GDP per Capita
```{r}
obs <- 1:nrow(gapminder)
for (i in obs) {
  gapminder[i, "gdp"] <- gapminder[i, "pop"] * gapminder[i, "gdpPercap"]
}
```
## Creating logged variables with a for loop
```{r}
for (i in obs) {
  gapminder[i, "log_gdpPercap"] <- log(gapminder[i, "gdpPercap"])
  gapminder[i,"log_pop"] <- log(gapminder[i,"pop"])
}
```
## Same as the below
```{r}
gapminder$log_pop2 <- log(gapminder$pop)
all(gapminder$log_pop2==gapminder$log_pop)
```
## When loops are useful
```{r}
years <- unique(gapminder$year)
for (i in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year==i], na.rm=T)
  print(paste0(i, ": ", mean_le))
}
continents <- unique(gapminder$continent)
for (i in continents) {
  mean_le_c <- mean(gapminder$lifeExp[gapminder$continent==i & gapminder$year==2007], na.rm=T)
  print(paste0(i, ": ", mean_le_c))
}
for (i in continents) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le_c <- mean(gapminder$lifeExp[gapminder$continent==i & gapminder$year==j], na.rm=T)
    print(paste0(j, ": ", mean_le_c))
  }
}
```
## Gap in life expectancy narrowed over time?
```{r}
for (i in continents) {
  print(paste0("Continent:" , i)) 
  for (j in years) {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent==i & gapminder$year==j], na.rm=T)
    print(paste0(j, ": ", sd_le))
  }
}
```
## Apply!
# apply(matrix, 1 = row or 2 = column, function)
```{r}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean)
```
##Lapply
```{r}
lapply(gapminder, mean)
```
##sapply with anonymous functions
#x is the iterator
```{r}
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year==x]))
```

## while loops
```{r}
i <- 1987
while (i <= 2002) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year==i])
  print(paste0(i, ": ", sd_le))
  i <- i+5
}
```
## If/else statements
```{r}
set.seed(1)
random_year <- sample(years,1)
random_year

if (random_year > 1977) {
  print(random_year)
} else {
  print("sorry, random year is less than 1977")
  }
```
```{r}
threshold <- 1987

for (i in unique(gapminder$year)) {
  if (i >=threshold) {
    tmp <- mean(gapminder$pop[gapminder$year==i])
    print(paste0(i, ": Mean population is ", tmp))
   } else {
       print(paste(i, "is less than", threshold))
   }
}

```
## Writing Functions
```{r}
get_values <-
  function(df, variable = "continent") {
    vals <- unique(df[[variable]])
    print(paste0(variable,": ", vals))
  }
get_values(gapminder, variable = "country")
```
#example
```{r}
report <- 
  function(df, variable) {
    for (i in unique(df$continent)) {
    var <- df[[variable]][df$continent == i]
    mean_le <- mean(var)
    median_le <- median(var)
    min_le <- min(var)
    max_le <- max(var)
    cat("\nContinent:", i, 
        "\nMean Life Expectancy:", mean_le,
        "\nMin Life Expectancy:", min_le,
        "\nMax Life Expectancy:", max_le,
        "\n")
    }
  }

report(gapminder, "lifeExp")
```

## Functions and loops
```{r}
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }

viz_lm(gapminder, "lifeExp", "gdpPercap", 1977)
```